---
title: "Introduction to the gower package"
author: "Mark van der Loo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the gower package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Gower's distance measure between records.

Gower's distance can be used to measure how different two records are. The records may contain combinations of logical, numerical, categorical or text data. The distance is always a number between 0 (identical) and 1 
(maximally dissimilar). Here is a link to the original publication.

Gower (1971) [A general coefficient of similarity and some of its properties.](http://venus.unive.it/romanaz/modstat_ba/gowdis.pdf)  _Biometrics_ **27** 857-874.

Basically, a similarity measure is computed for each pair of variables from two compared records. These similarity measures are then combined to an overal measure.


This package modifies Gower's original similarity measure in the following ways.

- In stead of the original similarity _S_, the package returns the distance _1-S_.
- The original paper does not mention the concept of `NA`. Missing variables are skipped when computing the distance.
- The original paper does not mention character data. These are treated as categorical data.

## Computing Gower's distance with the package

The function ```gower_dist``` computes pairwise-distances between records.

```{r}
library(gower)
dat1 <- iris[1:10,]
dat2 <- iris[5:15,]
gower_dist(dat1, dat2)
```
If one data frame has less records than the other, the shortest one is recycled over (just like when you're adding two vectors of unequal length)

```{r}
gower_dist(iris[1,], dat1)
```

It is possible to control how columns from the two data sets are paired for comparison using the `pair_x` and `pair_y` arguments. This comes in handy when similar colums have different names accross datasets.  By default, columns with matching names are paired. The behaviour is somewhat similar to that of base R's merge in that respect.

```{r}
dat1 <- dat2 <- iris[1:10,]
names(dat2) <- tolower(names(dat2))
gower_dist(dat1, dat2)
# tell gower_dist to match columns 1..5 in dat1 with column 1..5 in dat2
gower_dist(dat1, dat2, pair_y=1:5)
```

## Computing the top-n matches

The function `gower_topn` returns a list with two arrays.
```{r}
dat1 <- iris[1:10,]
L <- gower_topn(x=dat1, y=iris, n=3)
L
```

The first array is called \code{index}. Each column corresponds to one row of `x`. The entries of each column
index the top _n_ best matches of that row in x with rows in `y`. In the previous example, the best match of the first row of `dat1` is record number ```r L$index[1,1]``` (this should be obvious, since they are the same record). 
 
 The second array is called `distance` and it contains the corresponding distances.





## Parallelization

The underlying algorithm is implemented in C and parallelized using [OpenMP](http://www.openmp.org). OpenMP is available on most systems that can run R. Please see [this section](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#OpenMP-support) of the writing R extensions manual for up-to-details details on which systems are supported. At the time of writing, OSX is the only system not supporting OpenMP.

If OpenMP is not supported, the package will still work, the core algorithms will just not be parallelized.
